Multiscale Modeling of the Paradoxical Roles of the Immune System in EMT-Mediated Cancer

### 2019-03-08 Paper/Paper.tex
results edts

\documentclass[11pt, preprint]{article}
\usepackage{outline}
\usepackage{float}
\usepackage{graphicx}
\usepackage{color}
\usepackage{amsmath}
\usepackage{cite}
\graphicspath{ {Figs/} }

@@

% A multiscale model of EMT-mediated cancer reveals 
\title{Multiscale Modeling of the Paradoxical Roles of the Immune System in EMT-Mediated Cancer}
\author{Daniel R. Bergman$^{1}$,
Matthew Karikomi\,$^{1}$,
Qing Nie\,$^{1,2,*}$
and Adam L. MacLean\,$^{3,*}$
}
\affil{
  $^1$Department of Mathematics, University of California, Irvine, and \\
  $^2$Department of Cell and Developmental Biology, University of California, Irvine, Irvine, CA 92697, USA \\
  $^3$Department of Biological Sciences, University of Southern California, Los Angeles, CA 90089, USA \\
  $^*$Correspondence: macleana@usc.edu (ALM); qnie@uci.edu (QN).
}

@@

@@

The immune system affects tumorigenesis and cancer progression in myriad intricate ways. New breakthroughs in the field of immunotherapy are beginning to produce new therapies that have had a significant impact on patient health and survival \cite{pardoll2012blockade,restifo2012adoptive}. Beyond direct tumor-immune cell interactions, the immune system also modulates the tumor microenvironment (TME), and indeed, immune cells can be considered as a part of the TME: responding to local inflammatory signals, targeting mutated cells, and potentially eradicating tumor cells and restoring homeostasis in a local context \cite{de2006paradoxical}.

@@

Tumor cells also interact with the immune system by initiating a sequence of events that results in the activation of the adaptive immune response. This is accomplished by the presentation of antigens that innate immune cells recognize and carry to lymph nodes to activate T cells (and other components) \cite{schreiber11_cancer}. The tumor also engages in cellular processes that indirectly shape the TME, for example by releasing transforming growth factor beta (TGF-$\beta$), thus shifting the TME towards a tumor-supportive environment by enhancing the immunosuppression via Treg cell activation \cite{schreiber11_cancer}.

@@

The effects of the immune system on a tumor can be (very coarsely) summarized into two modules. Cytotoxic immune cells (such as natural killer (NK) cells and cytotoxic T cells (CTLs)) seek out and lyse tumor cells, whereas the regulatory branch of the immune system (which will below be accounted for by T regulatory cells (Tregs)), slows down the \tcr{effector functions} of these other cells. [REF]

@@

In addition, the immune system plays important regulatory roles in pre-cancerous tissue, where inflammation can increase the probability of tumorigenesis (and/or decrease the time to cancer), particularly in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}. However, recent work raises questions regarding the relationship between inflammation and cancer \cite{guo2017multiscale}. 

@@

Epithelial-to-mesenchymal transition (EMT) describes reversible process by which cells with an epithelial phenotype transition into cells with a mesenchymal phenotype.
While epithelial cells are in-part defined by their adhesion to neighbors, mesenchymal cells show much greater ranges of motility, and perhaps stem-like properties \cite{nieto2016emt}, although controversy regarding `stemness' and EMT remains \cite{nie18_stem, sha19_intermediate}.  
Recent work has shown that -- rather than being a binary process -- at least two stable intermediate states exist on the EMT axis \cite{hong2015ovol2, jolly15_coupling}. Ongoing investigations into the plasticity and stability of EMT overlap with discussions elsewhere, e.g. of discrete vs. continuous processes during cell differentiation \cite{moris16_transition}. Intermediate states have emerged as a central mechanisms by which cell fates (and the noise inherent within in them) can be controlled \cite{maclean18_exploring, ta16_controlling, rackauckas18_meanindependent}. 

@@

While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of both tumor initiation and progression \cite{nieto2016emt}. In addition, TGF-$\beta$ is a master regulator of EMT [REF], thus tying it to the signaling networks involved in tumor-mediated immune responses, since Tregs release TGF-$\beta$ upon arriving at the tumor site\cite{terry2017new}. Thus, even by considering only a single signaling factor, we find these three components: the tumor, the immune system, and the EMT process to be inextricably linked. It thus strikes us as a priority to derive models to understand how interactions between each of these three components affect cancer incidence and progression.

@@

Below we develop a multiscale model with this goal: to study interactions between the tumor, immune cells and EMT. Many

\tcr{I think this is where I would add which types of cancer most likely to be affected by EMT. But which? And what sources?}
In this way, the tumor shifts the balance of local environment along the EMT spectrum.

While there might be many other characteristics distinctive of mesenchymal cells that need further exploration, two qualities of mesenchymal cells are particularly relevant in the context of cancer and the immune system:
1) mesenchymal cells are less susceptible to immune clearance\cite{terry2017new} and
%
2) mesenchymal cells proliferate at a slower rate. \tcr{Ref}

% MODELS of cancer-immune 
%\cite{West19_JCOClinicalCancerInformatics}

*******summary of MIE and MGA (move later?)

The immune evasiveness of mesenchymal cells is well-established with even a functional cause for this phenomenon.
As a cell is being targeted by cytotoxic immune cells for clearance, a physical connection between the two cells must be created in order to precipitate the lysing event.
This immunological synapse, however, is dependent upon the surface proteins of the target cell, and in the case of mesenchymal cells there is preliminary evidence that these surface markers are down-regulated in such a way that the immune synapse is more difficult to form\cite{terry2017new}.
There are also other structural changes that take place in mesenchymal cells that could contribute to this refractory nature towards immune clearance.
Regardless, the evidence currently demonstrates a ``slipperiness'' of mesenchymal cells in regards to immune lysing.
We will refer to this as mesenchymal immune evasion (MIE).

The other facet of mesenchymal cells is their slower proliferation rate, and this is tied to the observation that mesenchymal cells often express a more stem-like phenotype.
Indeed, even in healthy tissue, this is readily observed and seen as a useful mechanism for homeostasis as epithelial cells can transition to mesenchymal, stem-like cells under appropriate conditions. \tcr{Ref}
While many qualities are associated with being stem-like, we choose to focus on a reduced proliferation rate. \tcr{Ref}
Cancer cells are well-known for achieving unregulated proliferation, so exploring a subset of cancer cells with decreased proliferation could provide intriguing new results.
In our model, we refer to this property as mesenchymal growth arrest, or MGA.

****thisis ?

Recent work on cancer-immune interactions has shown that the inflammatory conditions of the TME can play large and sometimes surprising roles in tumor growth.
For example, a mathematical model was developed to couple the inflammatory state of the TME with the cell cycles of the stem cells by increasing proliferation, decreasing apoptosis, and enhancing DNA damage that leads to mutations\cite{guo2017multiscale}.
The work shows that while inflammation does have an oncogenic effect, it can also have an onco-protective role as periods of mild inflammation are better at preventing cancer than periods of no inflammation in patients that have intermittent bouts of severe inflammation.
However, the aspect of the model that included the immune system is minimal and the connection with EMT is not included.
%However, the aspect of the model that included the immune system was minimal: regardless of the inflammatory environment, any cells with sufficient pathway mutations (2 in their model) had a fixed probability of being cleared by the metabolism-immune balance response. 
Here, we include a more robust immune module and explore how this immune module will improve our understanding of inflammatory effects on the tumor, and also include the epithelial-mesenchymal axis.
%In addition, we explore an extra component of Guo's stem cells: where they lie on the epithelial-mesenchymal axis.
Such additions will allow us to explore the interactions of the tumor, the immune system, and the epithelial-mesenchymal spectrum as well as the effects of the TME on these interactions.
%We will also study the effects of the TME on these interactions.

Next, we introduce the model modularly, which is appropriate given its multiscale and hybrid nature.
We then perform sensitivity analysis and identify parameters that are crucial for the progression to tumorigenesis.
We analyze the competing effects of EMT and of the immune system on tumorigenesis and find that EMT intricately regulates tumorigenesis: under certain regimes a careful balance of EMT- and immune-driven process can significantly prolong cancer-free survival.

@@

\section{Methods}
To address the questions around the relationships among cancer, the immune system, and EMT, we extended the multiscale, agent-based model \tcr{REF} to include an immune module and an EMT axis for the tissue cells.
The agents in the model are the tissue cells with tumorigenic potential.
They can either be mutation-free, or have any combination of three possible pathway mutations.
These three pathways are the three components within each cell in Figure \ref{fig:ModelIntro}A.
If the pathway is mutated, the component is colored red and has an `X'.
A single mutation is sufficient for the cell being counted as a mutant cell.

We model the immune cells as discrete variables but as one collective, in other words they are not agents themselves.
The cytokine TGF-$\beta$ is a continuous variables within the tumor microenvironment (TME).
It is worth noting that there is no spatial aspect to this model.
In doing this, we are assuming that cells are well-mixed as are all the relevant cytokines and other molecules acting on the system.
Finally, tissue cells can be labeled either epithelial or mesenchymal, and this labeling can change over time and depends both on the TME and the internal workings of each individual cell.
While the score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.
In Figure \ref{fig:ModelIntro}A, the EMT labeling is given by the color of the cell.
Even though each cell does have an EMT score on a spectrum of values, only the labeling influences its interactions in the model.

@@

\paragraph{Proliferation}
There are four factors that contribute to the probability that a cell will proliferate.
The first is a base proliferation probability that all cells have, $p$.
Second, if the cell has a mutation in the proliferation pathway, then the weight for proliferation is proportionally increased by $\text{prolif}_{\text{up}}$.
Third, if the cell is mesenchymal, then the weight for proliferation is proportionally decreased by $m_{GA}$, which stands for mesenchymal growth arrest.
Fourth, there is a negative feedback of the cells on their own proliferation which is quantified by a Hill factor as a function of the tissue cell population, $N$.
In total, the weight for proliferation is given by

@@

\subsubsection{Progression and tumorigenesis}
At the end of each cell cycle, the proportion of tissue cells that are mutated is calculated, and if it is above a certain threshold, a tumorigenic event is recorded and the patient is determined to have cancer.
The Time to Cancer is given as the time from the start of the simulation, not including warmup, until tumorigenesis.
Simulations run until either a tumorigenic event is recorded or until the maximum number of cycles has been reached.

@@

Once we had these parameter values, we created Gaussian distributions centered at these values with large variances and truncated at 0.
Some of the parameters were also truncated to the right because they represented a proportion, such as MIE and MGA, or because we simply wanted to keep the values within a reasonable range, such as the inflammation low and high duration parameters.
The variances were chosen as twice the mean to get a large spread of values.
%%%%%%%%%%%%%%%%%%%%%%%
%                          RESULTS                         %
%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{figure}[H]
\center
\frame{\includegraphics[width=\columnwidth]{Figure1/Figure1.pdf}}
\caption{A. Schematic depiction of the model with blue/red denoting non-mutated/mutated cells in the agent-based model. In each cycle, the fate of a cell is regulated by the EMT state of the cell and by any pathway mutations harbored. Inset shows major processes of immune system regulation on the tissue cells. The parameters used in this model can be found in the Supplementary material. Unless otherwise stated, parameter values are assumed to be those found in the Supplement. \tcr{Thoughts on the wording here? Once I have the table, I will use the name of that table.}
B. Example simulation of the dynamics of a one patient; warmup period is shown in red.
The inflammation cycling scheme is graphically represented above the patient dynamics.
C. Example survival curve for one cohort of patients with basal parameter values.
%A Should I list out all parameter values? A subset? Or refer to supplement for base parameter values?
D. Global sensitivity analysis of model parameters using the Morris one-step-at-a-time method.}
\label{fig:ModelIntro}
\end{figure}

\begin{figure}[H]
\center
\frame{\includegraphics[width=\columnwidth]{Figure2/Figure2.jpg}}
\caption{Global sensitivity analysis of model parameters using the Morris one-step-at-a-time method.}
\label{fig:MOAT}
\end{figure}

\subsection{A multiscale agent-based model of EMT-immune-tissue cell interactions and tumorigenesis}\label{ExplModel}
We begin by studying the model under baseline conditions to assess how its various components interact and effect the overal probability of tumorigenesis (Time to Cancer). In Figure \ref{fig:ModelIntro}A a schematic description of the model is shown. Within a single cell cycle, different choices are available to a cell and these are influenced by pathway mutations or EMT. For example, if a cell undergoes EMT, the probability that is will proliferate is reduced. Also shown are the different means by which the immune system acts on the tissue cells (\ref{fig:ModelIntro}A Inset). The NKs and CTLs attempt to clear out mutated tissue cells and deactivate upon successfully carrying out this cytotoxic function. The Tregs inhibit this cytotoxic activity. In addition, Tregs release TGF-$\beta$ which helps recruit more Tregs and also drives EMT.
%A \tcr{TGF-$\beta$ is [describe briefly action of each immune cell type]}  --Adam: Is this good?

For a typical {\it in silico} patient (see Figure \ref{fig:ModelIntro}B), after the warmup period, some cells mutate but are cleared before establishing a tumor. However, eventually one mutated cell is able to establish a growing tumor before being cleared. This happens around cell cycle 1000. At this point, the adaptive immune system (CTLs and Tregs) are being heavily recruited and TGF-$\beta$ is peaking. The number of mesenchymal cells grows with the increased TGF-$\beta$, but decreases rapidly after the tissue cells are entirely mutants. Interestingly, this return to a more epithelial phenotype occurs even as TGF-$\beta$ reaches its maximum value for the simulation.
%\tcr{[What happens??]}
The inflammation cycling scheme for this patient is shown at the top of Figure \ref{fig:ModelIntro}B and is repeated throughout, including during the warmup.
We will return to this shortly.

%\tcr{[What is?]} 
At cell cycle number 841 the proportion of mutated cells reaches 50\% and so the model marks the patient as having a Time to Cancer of 841, or 630.75 days. 
At the same time as mutated cells are taking over, many cells transition to a mesenchymal phenotype.
After the Time to Cancer, the mutated cells continue their rapid growth and soon make up 100\% of the cells.
Meanwhile, all the cells become mesenchymal for a short duration before most transition back to an epithelial state.

Considering the immune populations, we see that the NK population is approximately constant, while the lymphocyte populations grow quickly and dwarf the NK population following the accumulation of mutations.
The lymphocyte populations exhibit oscillatory behavior, however note that this is not due to intrinsic dynamics but rather due to the inflammation scheme that this patient is undergoing: 
the patient alternates between 30 cell cycles of high inflammation and 60 cell cycles of low inflammation.
In each of these periods periods, the lymphocyte populations increase or decrease rapidly in accordance with the inflammation.

We next consider a cohort of $N=500$ patients, and simulate the survival curve for these patients (Figure \ref{fig:ModelIntro}C). We see that all patients survive for approximately 100 cell cycles (75 days), before a time period where the cancer onset rate is roughly constant, between the approximate range of ($T = 100, T = 300$).

\subsection{Identification of key model parameters via Morris global sensitivity analysis}\label{SensAnalysis}
To assess the sensitivity responses \tcr{Is this the right phrasing? It sounds funny to me, but I see Adam chose this wording.} produced by the model with respect to the 31 model parameters, a Morris one-step-at-a-time (OAT) algorithm was implemented.
The results of Morris OAT sensitivity analysis are shown in Figure \ref{fig:MOAT}. A subset of parameters demonstrate much higher levels of sensitivity than others.
The two most influential according to this analysis are the death rate and the recruitment rate of Treg cells, this is most likely due to the dual roles Treg cells play in both suppressing the cytotoxic effects of other immune cells and secreting TGF-$\beta$, which drives EMT.
This ties Treg cells to all three components of the model.
Since below we seek to separate the effects of different model components, we do not choose the parameters influencing Treg cells for further analysis below.

Any of the parameters in the table which include the word ``low'' are parameters that govern the model in a low inflammatory state.
With the exception of ``prolif up'', all parameters that end in ``up'' or have `U' at the end of their abbreviation, are the factors by which these parameters change when the system switches from low to high inflammation.
The IHD and ILD parameters determine the duration of high and low inflammation, respectively.
A cell with mutated pathways will be affected by the ``apop down'', ``prolif up'', and ``immune evasion'' parameters, depending on which pathways are mutated.
The mesenchymal cells are influenced by the MGA and MIE parameters.
Finally, $\sigma$, $k_\text{EMT}$, TGFB$_{max}$, and $\gamma_\text{EC50}$ all control how cells transition between the epithelial and mesenchymal states.

We want to better isolate the effects of EMT on these dynamics, so parameters such as MIE and MGA are of interest.
In addition, inflammation parameters dictating the cycling scheme are important to study\tcr{REF}: we find that there are also high sensitivities in this model to these parameters.
In terms of Treg cells, their secretion of TGF-$\beta$ is highly significant and we will also study it further.

\subsection{Mesenchymal phenotypic properties dramatically change survival outcomes}\label{MesPars}
The two quantities that change when a cell transitions to a mesenchymal state are its immune evasion and its proliferation rate, MIE and MGA, respectively.
Both parameters are proportions between 0 and 1 with higher values indicating more extreme expression of mesenchymal properties.
MIE is the proportional reduction of the probability a mutated mesenchymal cell will undergo immune clearance in a given cycle.
MGA, which stands for mesenchymal growth arrest, is the proportional reduction in any mesenchymal cell of its chance to proliferate in a given cell cycle. 
The reduction in proliferation probability corresponds with an equal increase in probability for resting during that cell cycle.  
%\tcr{[briefly redefine what these terms do]}
Also involved in the EMT process, is the cytokine TGF-$\beta$, which upregulates EMT.
By the Morris OAT analysis presented in Section \ref{SensAnalysis}, these three parameters do have large impacts on survival outcomes and each warrant exploration.

We found that as the level of mesenchymal immune evasion (MIE) increases, Time to Cancer decreases ( Figure \ref{fig:FirstSurvivalCurves}A).
This result holds for all parameter sets studied, and indicates the simple relationship that mesenchymal immune evasion has with Time to Cancer.
As this subpopulation of the tumor becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
These results are summarized by Figure \ref{fig:FirstSurvivalCurves}A.

>>>>

\documentclass[11pt]{article}
\usepackage{amsmath, amssymb, graphicx, outline, float, cite, color}

\input{structure.tex} % Specifies the document structure and loads requires packages

\graphicspath{ {Figs/} }

@@

% A multiscale model of EMT-mediated cancer reveals 
\title{Multiscale Modeling of the Paradoxical Roles of the Immune System in EMT-Mediated Cancer}
\author{Daniel R. Bergman$^{1}$,
Matthew Karikomi\,$^{1}$,
Qing Nie\,$^{1,2,*}$
and Adam L. MacLean\,$^{3,*}$
}
\affil{
  $^1$Department of Mathematics, University of California, Irvine,  Irvine, CA 92697, USA \\
  $^2$Department of Cell and Developmental Biology, University of California, Irvine, Irvine, CA 92697, USA \\
  $^3$Department of Biological Sciences, University of Southern California, Los Angeles, CA 90089, USA \\
  $^*$Correspondence:  qnie@uci.edu (QN); macleana@usc.edu (ALM).
}

@@

\date{}

@@

Cancer and the immune system interact in myriad intricate ways. Beyond direct tumor-immune cell interactions, the immune system modulates the tumor microenvironment (TME), indeed, immune cells can be considered as a part of the TME: responding to local inflammatory signals, targeting mutated cells, and eradicating tumor cells to potentially restore homeostasis in a local context \cite{de2006paradoxical}. Recent breakthroughs in the field of immunotherapy are beginning to produce therapies that have had a significant impact on patient health and survival \cite{pardoll2012blockade,restifo2012adoptive}.

@@

Tumor cells interact with the immune system by initiating a sequence of events that results in the activation of the adaptive immune response. This is accomplished by the presentation of antigens recognized by innate immune cells that are transported  to lymph nodes where T cells (and other components) can be activated \cite{schreiber11_cancer}. The tumor also engages in cellular processes that indirectly shape the TME, for example by releasing transforming growth factor beta (TGF-$\beta$), thus shifting the TME towards a tumor-supportive environment by enhancing immunosuppression via Treg cell activation \cite{schreiber11_cancer}.

@@

The effects of the immune system on a tumor can be broadly summarized into two modules. The cytotoxic branch of the immune system (such as natural killer (NK) cells and cytotoxic T cells (CTLs)) seek out and lyse tumor cells. The regulatory branch of the immune system (T regulatory cells (Tregs) and other factors), inhibits the effective functioning of the cytotoxic branch. [REF]

@@

In addition, the immune system plays important regulatory roles in pre-cancerous tissue, where inflammation can increase the probability of tumorigenesis (and/or decrease the time to cancer), particularly in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}. Recent work raises some questions about the relationship between inflammation and cancer, suggesting that under certain conditions inflammation may not be oncogenic but rather onco-protective \cite{guo2017multiscale}. 

@@

Epithelial-to-mesenchymal transition (EMT) describes a reversible process by which cells with an epithelial phenotype transition into cells with a mesenchymal phenotype.
While epithelial cells are in part defined by their adhesion to neighbors, mesenchymal cells show much greater ranges of motility, and may display stem-like properties \cite{nieto2016emt}, although controversy regarding `stemness' and EMT remains \cite{nie18_stem, sha19_intermediate}.  
Recent work has shown that -- rather than being a binary process -- at least two stable intermediate states exist on the EMT axis \cite{hong2015ovol2, jolly15_coupling}. Ongoing investigations into the plasticity and stability of EMT overlap with discussions elsewhere, e.g. of discrete vs. continuous processes during cell differentiation \cite{moris16_transition}. Intermediate states have emerged as a central mechanism by which cell fates (and the noise inherent within in them) can be controlled \cite{maclean18_exploring, ta16_controlling, rackauckas18_meanindependent}. 

@@

While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of both tumor initiation and progression \cite{nieto2016emt}.  TGF-$\beta$ is a master regulator of EMT [REF], linking it to the signaling networks involved in tumor-mediated immune responses, since Tregs release TGF-$\beta$ upon arriving at the tumor site\cite{terry2017new}. Thus, even by considering only a single signaling factor, we find these three components (the tumor, the immune system, and EMT) to be linked. It therefore strikes us as a priority to develop models to understand how interactions between each of these three components affect cancer incidence and progression.

@@

Two features of the mesenchymal phenotype are of particular relevance in the context of cancer and the immune system: i) mesenchymal cells are less susceptible to immune clearance \cite{terry2017new}; and ii) mesenchymal cells proliferate on average at a slower rate than epithelial cells.
The immune evasiveness of mesenchymal cells is well-established: as a cell is targeted by cytotoxic immune cells for clearance, a physical connection between the two cells must be established before the lysing event.
This immunological synapse is dependent upon the surface proteins [WHICH??] of the target cell, and in the case of mesenchymal cells there is preliminary evidence that these surface markers are down-regulated in such a way that the immune synapse is more difficult to form \cite{terry2017new}.
%There are also other structural changes that take place in mesenchymal cells that could contribute to this refractory nature towards immune clearance.
%Regardless, the evidence currently demonstrates a ``slipperiness'' of mesenchymal cells in regards to immune lysing.
Below we refer to this phenotype as mesenchymal immune evasion (MIE).
The other facet of mesenchymal cells is their slower proliferation rate, and this is tied to the observation that mesenchymal cells often express a more stem-like phenotype.
Indeed, even in healthy tissue, this is readily observed and seen as a useful mechanism for homeostasis as epithelial cells can transition to mesenchymal, stem-like cells under appropriate conditions. \tcr{Ref}
While many qualities are associated with being stem-like, we choose to focus on a reduced proliferation rate. \tcr{Ref}
Cancer cells are well-known for achieving unregulated proliferation, so exploring a subset of cancer cells with decreased proliferation could provide intriguing new results.
In our model, we refer to this property as mesenchymal growth arrest, or MGA.
Mathematical oncology, that is, mathematical models of cancer incidence, progression, and treatment, has become a well-developed field; many models have offered insight into the cellular interactions underlying cancer and its interplay with the immune system, including older \cite{sherrattjonathana.92_oncogenes, pillis05_validated} and more recent works \cite{an15_agentbased, serre16_mathematical, louzoun14_mathematical, benzekry17_mathematical, west19_immune, owen11_mathematical-1, guo2017multiscale}. These studies have increased our understanding of how tumors grow in the presence of various immune components, and how treatment regimes can be designed to maximise the efficacy of cytotoxicity while minimizing risks to the patient. However, to our knowledge no models have addressed how the effects of EMT alter interactions between the immune system and cancer, and the subsequent implications for treatment. 
Here we develop a model with the goal of studying interactions between the tumor, the immune system and EMT. We seek to describe a set of crucial molecular and cellular interactions in (epithelial) tissue cells, including effects due to DNA damage and mutation, to investigate the probability that cancer will occur and, if so, when. A recent model of cancer-immune interactions \cite{guo2017multiscale} described the effects of the TME on the risk of cancer, and we build on the core cell cycle component of this model, adding significant new interactions to the immune component of the model (which was previously modeled by a single interaction), as well as the EMT component to the model. We seek to understand whether this more complex immune module will change our understanding of inflammatory effects on the tumor, and how the epithelial-mesenchymal axis influences these. In the next section we develop the model modularly to explain the intuition behind each component, and then go on to analyze its behavior. Global sensitivity analysis identifies parameters that are crucial for tumorigenesis. We then study these in more depth, focusing on the competing effects of EMT and of the immune system on tumorigenesis: we find that EMT intricately regulates tumorigenesis and that under certain regimes a careful balance of EMT- and immune-driven processes can significantly prolong cancer-free survival. To test these predictions, we analyze data from the Cancer Genome Atlas (TCGA), and find evidence for the synergistic effects predicted by the model for patients with pancreatic cancer.

@@

\section{Methods}
We develop an agent-based model to describe the relationships between cancer, the immune system, and EMT, building on cell-cycle and tissue-cell components of \cite{guo2017multiscale}. 
The agents in the model are the tissue cells with tumorigenic potential.
These can either be mutation-free, or, resulting from DNA damage during the cell cycle, can have any combination of three possible pathway mutations (shown in Fig. \ref{fig:ModelIntro}A).
If the pathway is mutated, the component is colored red and is marked with `X'.
A single mutation is sufficient for the cell to be considered as a mutant cell.

We model immune cells as continuous variables, appropriate since at the level of immune cells it is not necessary to keep track of single cell effects.
The cytokine TGF-$\beta$ is continuous within the tumor microenvironment (TME), i.e. the system is well-mixed.
Tissue cells can take on either epithelial or mesenchymal phenotypes in a plastic manner: these phenotypes depend on both on the TME and cell-intrinsic factors.
While the score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.
In Figure \ref{fig:ModelIntro}A, the EMT labeling is given by the color of the cell.
Even though each cell does have an EMT score on a spectrum of values, only the labeling influences its interactions in the model.

@@

\paragraph{Proliferation}

\tcr{In next few subsections the parameters are not defined. Need to define in text and/or make a big table to summarize all. On notation: some changes may be needed, e.g. $\delta_{MES}$ not very clear.}

There are four factors that contribute to the probability that a cell will proliferate.
The first is a base proliferation probability that all cells have, $p$.
Second, if the cell has a mutation in the proliferation pathway, then the weight for proliferation is proportionally increased by $\text{prolif}_{\text{up}}$.
Third, if the cell is mesenchymal, then the weight for proliferation is proportionally decreased by $m_{GA}$, which stands for mesenchymal growth arrest.
Fourth, there is a negative feedback of the cells on their own proliferation which is quantified by a Hill factor as a function of the tissue cell population, $N$.
In total, the weight for proliferation is given by

@@

\subsubsection{Mutation and tumorigenesis}
At the end of each cell cycle, the proportion of tissue cells that are mutated is calculated, and if it is above a certain threshold, a tumorigenic event is recorded and the patient is determined to have cancer.
The Time to Cancer is given as the time from the start of the simulation, not including warmup, until tumorigenesis.
Simulations run until either a tumorigenic event is recorded or until the maximum number of cycles has been reached.

@@

Once we had these parameter values, we created Gaussian distributions centered at these values with large variances and truncated at 0.
Some of the parameters were also truncated to the right because they represented a proportion, such as MIE and MGA, or because we simply wanted to keep the values within a reasonable range, such as the inflammation low and high duration parameters.
The variances were chosen as twice the mean to get a large spread of values.

%%%%%%%%%%%%%%%%%%%%%%%
%                          RESULTS                         %
%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{A multiscale agent-based model of EMT-immune-tissue cell interactions to study tumorigenesis}\label{ExplModel}

\begin{figure}
\center
{\includegraphics[width=0.8\textwidth]{Figure1/Figure1.pdf}}
\caption{A. Schematic depiction of the model with blue/red denoting non-mutated/mutated cells in the agent-based model. In each cycle, the fate of a cell is regulated by the EMT state of the cell and by any pathway mutations harbored. Inset shows major processes of immune system regulation on the tissue cells. The parameters used in this model can be found in the Supplementary material. Unless otherwise stated, parameter values are assumed to be those found in the Supplement. \tcr{Thoughts on the wording here? Once I have the table, I will use the name of that table.}
B. Example simulation of the dynamics of a one patient; warmup period is shown in red.
The inflammation cycling scheme is graphically represented above the patient dynamics.
C. Example survival curve for one cohort of patients with basal parameter values.
%A Should I list out all parameter values? A subset? Or refer to supplement for base parameter values?
D. Global sensitivity analysis of model parameters using the Morris one-step-at-a-time method.}
\label{fig:ModelIntro}
\end{figure}

We begin by investigating general features of the model to establish baseline conditions and to assess the impact of various model components on the key measured outcomes: the probability of cancer, and the Time to Cancer. 
Within the cell cycle, cell fate is determined via a set of rules that are influenced by EMT and immune interactions (Fig. \ref{fig:ModelIntro}A). For example, if a cell undergoes EMT, the probability that is will proliferate is reduced; if a cell gains a mutation in the apoptosis pathway, the probability that it undergoes apoptosis is greatly reduced. The different means by which the immune system acts on tissue cells are also shown (\ref{fig:ModelIntro}A Inset). NK cells and CTLs attempt to clear mutated tissue cells, and deactivate upon successfully carrying out their cytotoxic function. T regulatory cells (Tregs) inhibit cytotoxic activity. Tregs also release TGF-$\beta$ which promotes the recruitment of Tregs, as well as increasing the rate of EMT.
\par
The inflammation cycling scheme for a typical {\it in silico} patient consists of alternating high and low regimes (Fig. \ref{fig:ModelIntro}B); the inflammation schemes modeled will be discussed in detail in the next section. For this patient, after the warmup period, cell mutations are observed at a rate low enough that they are cleared by cytotoxic cells before being able to establish a tumor. However, after approximately 1000 cell cycles (750 days), one mutated cell exhibits clonal growth, and a tumor is established. At this point, large numbers of cells from the adaptive immune system (CTLs and Tregs) are being heavily recruited, and a peak in TGF-$\beta$ expression is observed. The number of mesenchymal cells increases initially with increasing TGF-$\beta$, but decreases after the clone takes over the tissue. This shifts back towards an epithelial tissue phenotype occurs even though the expression of TGF-$\beta$ remains high.
After 841 cell cycles, the proportion of mutated cells reaches 50\%: this is modeled as the threshold defining tumorigenesis. Thus this patient has a Time to Cancer of 841 cell cycles, or 631 days. 
After cancer incidence, the number of mutated cells continues to grow rapidly and soon makes up 100\% of the cell population. A peak in mesenchymal phenotypes is observed following cancer incidence, before most transition back to an epithelial state.
\par 
Considering the immune system dynamics, we see that the NK population is approximately constant, while the adaptive populations (CTL and Treg cells) grow quickly and dwarf NK cells in number following the accumulation of oncogenic mutations.
%Note: NK cells also lymphocytes! 
The adaptive immune populations also exhibit oscillatory behavior, however note that this is not due to intrinsic dynamics but rather due to the inflammation scheme that the patient is undergoing: alternating between 30 cell cycles of high inflammation and 60 cell cycles of low inflammation. Within each of these periods, adaptive immune populations increase or decrease rapidly in accordance with the inflammation state.
\par 
The {\em in silico} patient described here is given for the purpose of illustrating features the model. Given the multiple sources of noise in the model, in order to quantify patient dynamics and cancer-free survival rates, below we will simulate large cohorts of patients. In Fig. \ref{fig:ModelIntro}C we simulate the survival curve for a cohort of $500$ patients: we see that all the patients survive for approximately 100 cell cycles (75 days). Subsequently, in the approximate range of $T= [100,300]$, the cancer onset rate is roughly constant, and after $600$ cell cycles, no patients remain cancer free.

\subsection{Identification of key regulatory parameters via Morris global sensitivity analysis}\label{SensAnalysis}
To assess the sensitivity of the model parameters and thus identify those that are most important in determining cancer-free survival times, we applied Morris one-step-at-a-time (OAT) sensitivity analysis (see Methods) to the model.
% Sentence on how it is 'global' and sentence on what the measured output is (ie y axis) 
...
The results of Morris OAT on the 31 model parameters are shown in \ref{fig:MOAT}: a subset of parameters demonstrate much higher levels of sensitivity than others.
The two most influential according to this analysis are the death rate and the recruitment rate of Treg cells, this is most likely due to the dual roles Treg cells play in both suppressing the cytotoxic effects of other immune cells and secreting TGF-$\beta$, which drives EMT.
This ties Treg cells to all three components of the model.
Since we seek to separate the effects of different model components, we do not choose the parameters influencing Treg cells for detailed analysis below.
\par
%% We need to discuss parameter names. Maybe replace 'low' with 'base' and 'up' with 'gain' ?
Any of the parameters in the \tcr{table} which include the word ``low'' are parameters that govern the model in a low inflammatory state.
With the exception of ``prolif up'', all parameters that end in ``up'' or have `U' at the end of their abbreviation, are the factors by which these parameters change when the system switches from low to high inflammation.
The IHD and ILD parameters determine the duration of high and low inflammation, respectively.
A cell with mutated pathways will be affected by the ``apop down'', ``prolif up'', and ``immune evasion'' parameters, depending on which pathways are mutated.
The mesenchymal cells are influenced by the MGA and MIE parameters.
Finally, $\sigma$, $k_\text{EMT}$, TGFB$_{max}$, and $\gamma_\text{EC50}$ all control how cells transition between the epithelial and mesenchymal states.

Since one goal of our analysis is to assess the specific effects of EMT on immune-cancer dynamics, the parameters MIE and MGA are of particular interest.
In addition, inflammation parameters dictating the cycling scheme are important to study\tcr{REF}: we find that there are also high sensitivities in this model to these parameters.
In terms of Treg cells, their secretion of TGF-$\beta$ is highly significant and we will also study it further.

\begin{figure}
\center
{\includegraphics[width=0.85\textwidth]{Figure2/Figure2.jpg}}
\caption{Global sensitivity analysis of model parameters using the Morris one-step-at-a-time method.}
\label{fig:MOAT}
\end{figure}

\subsection{Mesenchymal phenotypic properties dramatically change cancer-free survival times}\label{MesPars}

\begin{figure}
\center
{\includegraphics[width=0.85\textwidth]{Figure3/Figure3.pdf}}
\caption{
A-C. Trajectory of one patient per cohort followed from warmup through 2000 cell cycles while varying MIE (A), MGA (B) or XXXXXX (C). 
D. Increasing MIE decreases Time to Cancer. 
E. Increasing MGA increases Time to Cancer.
F. Increasing Treg TGF-$\beta$ production results in decreased Time to Cancer.
%% \color{red} Cox regression or KM test statistics to be computed to demonstrate the significance of this claim.
}
\label{fig:FirstSurvivalCurves}
\end{figure}
